To be FAIR: Theory Specification Needs an Update

### 2025-06-08 test.Rmd
update 2025-06-08 14:18:50.386429

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers^[[As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.]{#faircap}].
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
[FAIR theory has the potential to improve the efficiency of scholarly communication and]{#test2}
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory. 
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical processes of theory construction and improvement have been largely overlooked.
<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences. Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].  NvD dissents: -->
<!-- NvD: I cannot accept this footnote as an adequate characterization. Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more). See for instance Brian Haig's 2005 paper. Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions". At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree. I am fine with the last sentence though ;-) My suggestion is to write the following: -->
<!-- NvD: We use induction to describe inferences from specific observations to general theories. In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations. We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable. -->
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the *empirical cycle* [@degrootMethodologyFoundationsInference1969],
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases, see Figure \@ref(fig:figecs)a.
In one half of the cycle, labeled the "Context of Justification" by Wagenmakers and colleagues, hypotheses derived from theory are tested on data. In the other half of the cycle (the "Context of Discovery"),
patterns observed in data are generalized to theoretical principles, Figure \@ref(fig:figecs)b.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

@@

Merely publishing theory in a journal article does not make it open;
to be open, theory should adhere to established open science standards for specification and archival.
We propose to implement theories as information artifacts,
and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data". 
They provide information about the nature and content of an information artifact and are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g. also such things as scientific practice (experimental design, materials)). -->
<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added). -->
<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives. We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated. But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory. We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.  -->
Definitions of theory are abundant, and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our suggested approach is not limited to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p. 336].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
Over time, more precise *statistical/mathematical models* could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Going even further, a *generative/computational model* could be specified,
which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
More rigorous formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) vagueness.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, formal theories are not automatically FAIR.
The FAIR principles
<!-- pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility. -->
<!-- They  -->
can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).

@@

We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
These resources are often merely described in papers (if at all), and not made available for reuse.
Modular publishing is the practice of making each of these resources available as independent *information artifacts*,
facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].</span>
<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>

Modular publishing can be achieved by archiving specific resources (including theory) in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts.
<!-- - as does, for example, the [Crossref](https://www.crossref.org/) association, -->
<!-- which is used by many academic publishers. -->
The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, data can be archived in Zenodo with the metadata property `resourceType: dataset`.
If the data were collected for a specific paper, that relationship can be cross-referenced with `relationType: IsSupplementTo`.
Similarly, a FAIR theory can be connected to a specific paper by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
Other cross-references are useful for documenting relationships between multiple theory objects: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<span id="modularpublisingdetract">
Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for greater nuance and the author's voice.
Theories published in traditional papers can be supplemented by FAIR versions that live independently, evolve collaboratively, and feed into reproducible workflows.
</span>

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
Note that archival of the version of record is *not* a function of Git(Hub).
While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.

@@

Aside from technical solutions, version control is a social process as well.
On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
<!-- NvD: Maybe we can add an  example (e.g., the theory that explains ego-depletion and the failed replication of the phenomenon). -->
In computer science, these challenges are navigated by assigning version numbers.
Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to information artifacts.
Whereas version control tracks changes,
semantic versioning communicates what those changes mean to users of the theory,
guides the social process of theory development, and signals how much a theory has been changed.
We propose the following adaptation of semantic versioning for theories:

@@

The four Findability criteria are applicable to theory with only minor adjustments, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artifacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena.
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

@@

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate information artifacts.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the information artifact [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as a centralized repository for information artifacts, researchers may begin to search there more regularly.
Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016, p. 2].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal (an example of such a DOT representation is given below).

>>>>

<span id="whatsfair">The FAIR Guiding Principles (hereafter: FAIR principles) were established by a diverse consortium of stakeholders to improve the reusability of research data and other resources produced during the course of scholarly work
by making them Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016]^[[As the colloquial use of these terms differs from their definition according to the FAIR principles, we capitalize these terms when referring to specific FAIR principles.]{#faircap}].
Since the FAIR principles' inception, 
they have become a widely adopted standard for archival of academic output,
representing an estimated tens of billions of dollars in reuse value [@vogtFAIR20Extending2024].
Scholars have demonstrated their relevance for making other digital objects more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].</span>
The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially connected to - the scientific paper.
<span id="alldefinitions">Definitions of theory are abundant and hotly debated,
but many of them are compatible with the FAIR principles.</span>
FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One frequently raised concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
When theories are ambiguous,
precise predictions cannot be derived from them without
resorting to subjective interpretation or invoking additional assumptions,
which makes them harder to falsify.
A second concern has received less attention,
is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical processes of theory construction and improvement have been largely overlooked.
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

Given that a pluriformity of definitions are consistent with FAIR theory principles,
we do not limited ourselves to any one particular definition - although, at times, our writing will inevitably reveal a particular vantage point.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain (Table \@ref(tab:tabmeehl)):
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p. 336].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.

@@

According to the *empirical cycle* [@degrootMethodologyFoundationsInference1969],
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases, see Figure \@ref(fig:figecs)a.
In one half of the cycle, labeled the "Context of Justification" by Wagenmakers and colleagues, hypotheses derived from theory are tested on data [-@wagenmakersCreativityVerificationCyclePsychological2018]. In the other half of the cycle (the "Context of Discovery"),
patterns observed in data are generalized to theoretical principles, Figure \@ref(fig:figecs)b.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

@@

Merely publishing theory in a journal article does not make it open;
to be open, theory should adhere to established open science standards for specification and archival.
We propose to implement theories as digital objects,
and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data". 
They provide information about the nature and content of a digital object and are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues [-@lamprechtFAIRPrinciplesResearch2019s],
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary (minor) changes,
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR digital object,
and we provide worked examples to encourage their adoption.

@@

### What to Archive?

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be archived as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind of information is useful.
As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
Over time, more precise *statistical/mathematical models* could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Going even further, a *generative/computational model* could be specified,
which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).

@@

Concerns about the state of psychological theory have motivated increasing calls for greater theory "formalization" [@smaldinoModelsAreStupid2017; cf. @oudemaatmanPsychologysTheoryCrisis2021].
Formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) vagueness.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

Crucially for the present paper, the issue of formalization is orthogonal to FAIRification.
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it pertains to rigorous and transparent archival and communication about theories, with the aim of enhancing their reusability.
FAIR theory does not require formalization, and formal theories are not automatically FAIR.
The FAIR principles
apply to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).

@@

We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
These resources are often merely described in papers (if at all), and not made available for reuse.
Modular publishing is the practice of making each of these resources available as independent *digital objects*,
facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].</span>
<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>

Modular publishing can be achieved by archiving specific resources (including theory) in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for digital objects.
<!-- - as does, for example, the [Crossref](https://www.crossref.org/) association, -->
<!-- which is used by many academic publishers. -->
The DataCite Metadata Schema offers a standard way to document digital objects with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, data can be archived in Zenodo with the metadata property `resourceType: dataset`.
If the data were collected for a specific paper, that relationship can be cross-referenced with `relationType: IsSupplementTo`.
Similarly, a FAIR theory can be connected to a specific paper by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
Other cross-references are useful for documenting relationships between multiple theory objects: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<span id="modularpublisingdetract">
Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for greater nuance and the author's voice.
Theories published in traditional papers can be supplemented by FAIR versions that live independently, evolve collaboratively, and feed into reproducible workflows.
</span>

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving digital objects.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
Note that archival of the version of record is *not* a function of Git(Hub).
While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.

@@

Aside from technical solutions, version control is a social process as well.
On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
<!-- NvD: Maybe we can add an  example (e.g., the theory that explains ego-depletion and the failed replication of the phenomenon). -->
In computer science, these challenges are navigated by assigning version numbers.
Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to digital objects.
Whereas version control tracks changes,
semantic versioning communicates what those changes mean to users of the theory,
guides the social process of theory development, and signals how much a theory has been changed.
We propose the following adaptation of semantic versioning for theories:

@@

The four Findability criteria are applicable to theory with only minor adjustments, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for archived objects,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena.
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for a digital object which subsumes the DOIs of that artifact's versions.

@@

Findability is substantially amplified if intended users of a resource know where to search for it.
This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
Regrettably, most academic search engines only index traditional print publications, not other digital objects.
Since the status quo is to publish theories in papers,
the FAIR requirements are met if scholars continue to do so,
and additionally publish theories as separate digital objects.
The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
<!-- An ad-hoc solution is to publish a paper as documentation for the digital object [@mcgillivrayDeepImpactStudy2022]. -->
<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
<!-- The same solution could be applied to theories. -->
In the longer term, it may not be necessary to write a paper for each theory.
If Zenodo becomes more recognized as a centralized repository for digital objects, researchers may begin to search there more regularly.
Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index digital objects stored in Zenodo.

@@

Interoperability pertains to the property of digital objects to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016, p. 2].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all digital objects created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal (an example of such a DOT representation is given below).
